Query Optimization for Dynamic Imputation

### 2017-05-15 paper/rebuttal.tex

\section{Overall changes}
\begin{itemize}[leftmargin=*]
\item Redundant figures, excessive whitespace, formatting.

  \resp{Several reviewers commented on the amount of whitespace, so we've compressed many of the figures and removed whitespace to make room for additional material.}

\item \resp{
    We have improved our system since the paper was submitted by adding imputation for categorical variables.
    This has a minor impact on the experimental results.
    It improves the ability of the regression tree model to perform imputation by making these variables available to the model.
  }

\item \resp{We extended the original paper two additional popular imputation techniques: mean value imputation and hot-deck imputation.

We added a section discussing how to construct the imputation-related functions used by our cost model, which represent the API for a new imputation operator,
for these two new techniques and the reference technique used in the original submission (\Cref{sec:imputation})
}

\item \resp{We extended experiments to include the new operators and provide more details on results for $\alpha$ (\Cref{sec:results}). We also added the Pareto frontiers for various plans (\Cref{fig:pareto-frontiers} ) and
explored planning times for more complex queries.}

\item \resp{We included a new section that provides a roadmap for implementing \ProjectName{} on top of an existing database. (\Cref{sec:implementation})}
\end{itemize}

@@

    \resp{
      The outsized error in Q8 is due to a drop operator placed very early on in the plan-- this is a result of optimizing for performance
      and the fact that this data set has many missing values for relevant attributes (77.50\% for \texttt{studentdebtowe}). This produced significantly smaller sample sizes for various
      groups, which skewed the average calculation for each group significantly. For example, the average student debt for subjects with no high school degree was almost 83\% off from the estimate
      using the base table imputation, but if we look at the underlying population the former was calculated with 6 people, while the latter was with 258 people.

      We can trivially extend \ProjectName{} to warn users when the query chosen for execution has a high $\textsc{Penalty}$ estimate, along with the number of tuples that have
      been dropped (after execution), so that situations such as those observed in Q8 can be identified by the user. We've added this detail to the paper (\Cref{sec:results}).
    }

  \item Although it doesn't consider cost or the drop operator, a similar technique appears in Yang et. al. "Lenses: An On-Demand Approach to ETL." A comparison would be nice.

   \resp{
     Yang's work is definitely related, and we've added it to our related work section.
   Their system helps an analyst prioritize cleaning tasks for a particular query.
   It uses pre-trained regression models as one of a number of repair strategies for missing data rather than creating a query plan which incorporates the training and running of such models. (\Cref{subsec:missing-value})
   }

@@

  \item S2: Why no unions?

    \resp{We skipped unions for brevity. There is nothing fundamental about our algorithm which makes unions difficult to implement.}

  \item S3.4.1, Definition 2: On my first read-through, I was puzzled about why there was no plan with zero-loss. It would help to clarify that Loss is not about tuple loss, but rather a heuristic penalty function.

    \resp{We have clarified this in the paper (\Cref{sec:cost-model}) by emphasizing that \textit{Loss} is not a loss in the statistical sense or in the tuple loss sense and by renaming it to $\textsc{Penalty}$.}

@@

    \resp{This is a reasonable concern. Though often an appropriate proxy, optimizing for the maximum training set is flawed when the query requires the construction a large, skewed data set. The problem seems to be particular to queries which generate large intermediate tables. In some cases, as when joining two datasets, the intermediate table is a good target for training because it exposes correlations which would be hidden in the separate tables. In other cases, this intermediate table is large and skewed, which causes problems for imputation.

      Avoiding this case requires us to estimate the skew in these intermediate tables, which is an area that traditional query optimizers struggle with as well.
      However, if we had improved intermediate estimates, we could incorporate them to improve our system.
      }
    %\todo{Need a better explanation here.}

  \item S3.4.4: It appears that Time() assumes that plans are IO-bound. This is a perfectly reasonable assumption, but should be stated explicitly, as datasets that fit in memory will require a different cost model.

    \resp{Agreed. Our database stores tables in files, so our plans are I/O bound, but a database which uses an in-memory representation will need a different Time() function. We have clarified this in the paper (\Cref{sec:query-runtime}).}

@@

  \item S3.4.4: "Drop is a special case of a sequential scan, and its time complexity can be expressed as a function of the number of heap pages read and the I/O cost per page." seems to suggest that drop is implemented as a blocking operator. Why not just take the same approach as selection and treat it as having negligible cost?

    \resp{We mistakenly suggest in S3 that we treat \textit{Drop} as a blocking operator. In
      our implementation, it has $0$ cost, and we have updated the text to reflect that.}

  \item S4: All of the workload queries are (data-size-independent) AVG queries. What about other types of aggregates (e.g., SUM, COUNT)?

    \resp{Though we do not show this explicitly, we repeat all of the queries using COUNT
      aggregates in order to compute the ``Count Fraction''. The timing results are
      identical in this case and the number of tuples returned can be seen
      directly in Table 4. We would expect benchmark queries that use aggregates such as SUM 
      to work well in our system, with errors proportional to the fraction of relevant tuples retrieved
      and the AVG value estimated. We have noted this in the experiments section now (\Cref{sec:experiments}).}

@@

  \item S4.4.1: Imputation over the raw data is being used as ground truth. Is there any reason to suspect that this is the correct thing to do? Perhaps joining multiple tables together exposes new attributes that makes imputation *more* reliable? At the very least, it would be useful to understand why there is so much error in $\alpha=0$ on Q2 and Q8.

    \resp{For our purposes, imputation over the raw data is a reasonable ground truth. This
      is because the analyst, in practice, should be considering the tradeoffs on quality
      and performance of using ImputeDB versus a production imputation system on the raw
      data. Unfortunately, they would never have access to real ground truth in such a
      scenario, and moreover, finding ground truth for use even in an empirical evaluation
      of imputation methods can be challenging because artificially dirtying the data
      often does not capture the complex reasons for missingness in real world datasets.
      We updated the discussion in \Cref{sec:results} to emphasize this point.}

@@

  \item D1) The model relies on external imputation techniques, and given it's done on-the-fly, the imputed values will differ based on the qualifying query records?

    \resp{Yes, as discussed in the response to 1.}

  \item D2) The paper does not propose new imputation techniques but rely on existing statistical imputation algorithms. There is no comparative evaluation, nor samples of imputed values to show the goodness of the imputed values. It would be nice to see a comparative evaluation of different statistical imputation methods, and suggestions for extensions that can fit within the proposed imputation model.

    \resp{
      We have added two new imputation methods to the experiments (\Cref{sec:results}), to show that the planning algorithm can be used successfully with different imputation methods.
      We have also added a subsection to the algorithm section explaining how we chose the cost functions for each imputation method that we test and giving guidelines for choosing cost functions for a new imputation method (\Cref{sec:quality} and \Cref{sec:query-runtime}).
    }
  \end{itemize}

@@

  \item D1. Section 3, which is the meat of the paper, is very overloaded, without getting in depth in various parts of the system. It definitely should be split in more sections, and more details should be given. In particular:
    -- Give some more information about cardinality estimation. Also, Algorithm 3 is too far from the text, and needs some explanation too.
    -- Some more details are also required in the Imputation quality section. Why are you choosing these specific formulas for Loss, especially for the impute operator. Some examples would help.
    -- Algorithms 1 and 2 could probably be presented in an easier to grasp way.

    \resp{
    We've broken up this section and consolidated algorithm notation. The cardinality estimation section provides more detail about the algorithm. We also
    added intuition for how the heuristic Penalty (a renaming of Loss, for clarity) and Time should be designed. We walk through these definitions for three different
    impute operators.
    }

  \item D2. It would be nice to show how exactly your approach can be implemented in an existing optimizer. This comes up in various places in the paper:
    -- When defining the search space, why not first define the transformation operators that inject the imputation operators. Then any optimizer can pick these rules up and add them in its plan enumeration. Of course, the used approach that follows System-R's plan enumeration is useful, but why lose generality when presenting the search space. Moreover, showing the transformation rules can help you show the size of the search space that will better justify the restrictions you apply.
    -- Why use simplified cost formulas for the traditional relational operators and not existing systems' formulas?
    -- It would be great (but not necessary of course) to add your techniques in an existing DBMS, such as Postgres to see the impact on a real system in your experiments.

   \resp{
   We've added a section (\Cref{sec:implementation}) that describes in detail the components that need to be added to an existing database system to implement our technique. 
   The details come directly from our implementation of \ProjectName{} on top of SimpleDB, a popular teaching database used at MIT, University of Washington, and Northwestern University, 
   amongst others. These extensions could further exploit any existing sophistication in the underlying system for tasks such as execution cost estimates.

@@

  \item D4. How would a two-step approach that first finds the optimal plan (w/o imputation) and then adds the imputation operators perform?

    \resp{This approach would not be able to improve on the plans produced by our approach, as it would produce a subset of the plans that we consider. It's unclear how much worse the plans would be, but a major issue is that the Drop and Impute operators have an impact on the cardinality of their outputs. Our optimizer can use this information to improve the plan, but adding imputation after the fact would not be able to rearrange the plan to capitalize on changes in cardinality.}

  \item D5. Can you have plans with a mix of drop and impute operators?

    \resp{Yes, we have now highlighted this for clarity in the paper as well
    (\Cref{fig:example-plans} and \Cref{sec:search-space})}

@@

  \item D6. During plan enumeration, can you safely prune plans at some point, similar to what a traditional optimizer would do?

    \resp{We maintain the set of Pareto optimal plans at each point in the optimization. This set is significantly smaller than the set of all plans, so it represents some pruning (a point we have
    now clarified in the paper in \Cref{subsubsec:pareto}). We found that the size of these Pareto sets is small enough that we did not need to prune further.

@@

      If necessary, it is possible to sample from the Pareto optimal plans during the plan enumeration. The final selected plan may not be Pareto optimal, but this is a potential way to control the number of plans.
      To perform this approximation, we only add plans to the frontier that are non-dominated and are sufficiently different from existing plans in the frontier. The plan already in the frontier should
      provide similar tradeoffs and reduces the search space explored. \ProjectName{} only employs this optimization when a large number of tables requiring imputation are involved in the query.
      }

    \item D7. When defining the cost of impute operator, how do you know a priori the number of iterations for the algorithm (given it has to converge)?

  \resp{For the iterative regression impute operator, we set a predefined number of iterations as an upper bound on its execution. If the imputation converges,
  we can return immediately, otherwise we return after this upper bound is reached. This is a common approach used in iterative optimization techniques, such
  as gradient descent. We've clarified this technique in the section discussing this imputation operation (\Cref{subsubsec:decision-trees}).
}

@@

  \item D8. Are there impute implementation that are non-blocking? Or do they always have to perform multiple passes over the data?

    \resp{Yes. For example, in time series data, it is reasonable to fill the most recent non-null value forward. This kind of imputation isn't useful for survey data, so we didn't consider it for our experiments. We have extended the experiments section to consider a simple non-blocking imputation method relevant to our dataset: mean value imputation. This is a common technique, which fills in  missing values
    with the average observed value. In our case, we implement a non-blocking variant that estimates the mean from the plan's histograms, avoiding the need to perform a full pass over
    the data before imputing (\Cref{subsubsec:mean}).}

@@

  \item D9. Additional experiments I suggest to include:
    -- Consider a scenario where the user submits multiple similar queries. Show after how many such queries your approach has no longer a competitive advantage compared to performing imputation on the data.
    -- Try different values for $\alpha$ (not just 0 and 1).
    -- Show runtime of enumeration algorithm as queries get more complex.

    \resp{
      It is straightforward to compute the number of queries before we lose our advantage.
      For example, performing full imputation on the ACS dataset takes 355 minutes (Sec. 5.3.2) and we can run a simple query in 4 seconds, so we can run approximately $5,000$ queries before taking more time than full imputation. We've made this point explicit in the ACS experiment section now.

      The sensitivity to different values of $\alpha$ depends on the query itself. Performing a scan of values of $\alpha$ with small step sizes can return
      a lot of the same plans for execution. We did run experiments with a range of $\alpha$ values, but instead of presenting a figure with each value of $\alpha$,
      which can result in many duplicate plans (since different $\alpha$ values may result in the same plan), we extended the results section to include a subsection that presents the final frontiers for
      two queries, one from each dataset used during evaluation. This frontier can be exposed to the user to help in their workflow.

     The current set of results have been extended to show an additional intermediate $\alpha$ value.

@@

  \item D11. Less important points:
    -- Sections 4.2 and 4.3 in the experiments are too verbose.
    -- In Section 3.2, you mention that joins are performed after filtering. When filters are on results of joins, this is not possible.

    \resp{The subset of SQL that we consider does not allow filtering on the results of joins (see Section 3.2). Extending the planner to cover this case is not complex. Before adding filters, it must add imputations on the required attributes.}

  \item Given you optimize for both performance and quality, Tummer's and Koch's work on multi-objective QO is related.

    \resp{Yes, Tummer \& Koch is definitely related.
      However, much of the complexity of their method is in dealing with cost functions which are not well behaved.
      Our cost functions are monotonic, which allows us to use a simpler optimizer than what Tummer \& Koch propose.
    We have added a discussion of their work to the related works section (\Cref{sec:relatedwork}).}

### 2017-05-15 paper/algorithm.tex
paper: minor tweaks to alg. section

\section{Algorithm}
In order to correctly plan around missing values, we first extend the set of relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). 
The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
quality of the imputations performed and the runtime performance. Our imputation planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), in practice, planning times are not a concern for most queries.

@@

The space of query plans is similar to that considered in a canonical
cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
\Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.

@@

\subsubsection{Imputation quality} 
\label{sec:quality}
Every instantiation of the impute-operator defines a custom penalty function $P$ for imputing \nullv{} fields.
\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$, computed as follows.

@@

The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should take into account the complexity of the imputation algorithm.
Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of estimates.

We note here that for every imputation operator inserted into the query plan, a new
statistical model is instantiated and trained. Though it is tempting to try to pre-train
imputation models while the analyst begins to make queries, the exact rows of data, set
of complete attributes, and set of attributes to impute are unknown until runtime, making
this strategy infeasible. Thus, if applicable, $T$ should take into account both
training and application of the imputation model.

@@

\subsubsection{Planning algorithm}
The planner (\Cref{algo:top-level-planner}) operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}--\Cref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections
are pushed to the leaves, and
a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e.\ the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

The join order and imputation optimization is based on the algorithm used in System R~\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the Pareto frontiers.
The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.

>>>>

\section{Algorithm}
In order to correctly plan around missing values, we first extend the set of relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). 
The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates of the 
quality of the imputations performed and the runtime performance. Our planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}).
Finally, we show that while our algorithm is exponential in the number of joins (\Cref{sec:complexity}), planning times are not a concern in practice.

@@

The space of query plans is similar to that considered in a canonical cost-based optimizer~\cite{blasgen1981system}, with the addition of imputation operators.
\Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.

@@

\subsubsection{Imputation quality} 
\label{sec:quality}
Each imputation method has an associated cost function $P$, which is a measure of the quality of its output.
\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$, computed as follows.

@@

The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should take into account the complexity of the imputation algorithm.
Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of the estimates.

Note that for every imputation operator inserted into the query plan, a new
statistical model is instantiated and trained.
Though it is tempting to try to pre-train imputation models, the working data, set of complete attributes, and set of attributes to impute are unknown until runtime, making this strategy infeasible.
Thus, if applicable, $T$ should take into account both
training and application of the imputation model.

@@

\subsubsection{Planning algorithm}
The planner (\Cref{algo:top-level-planner}) operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}--\Cref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Next, it constructs a plan cache. For each table used in the query, any selections are pushed to the leaves, and a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}).
If no selections are available, a 
simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is then extended for any necessary grouping and aggregation
operations (\Cref{lst:line:group}) and for projections (\Cref{lst:line:proj}). The planner now contains the plan frontier (i.e.\ the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

The join order and imputation optimization is based on the algorithm used in System R~\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the Pareto frontiers.
